A Plasma Torus Around a Young Low-Mass Star

### 2025-03-19 paper/ms.tex
update

\begin{abstract}
\normalfont
Roughly one percent of red dwarfs younger than 100 million years
show structured, periodic optical light curves suggestive of
transiting opaque material that corotates with the star
\cite{Rebull2016,Stauffer2017,Rebull2018,Bouma2024}.  The
composition, origin, and even the existence of this material are
uncertain. The main alternative hypothesis is that these complex
periodic variables (CPVs) are explained by complex distributions of
bright or dark regions on the stellar surfaces
\cite{Koen2021}.  Here, we present time-series spectroscopy and
photometry of a rapidly-rotating ($P$=3.9\,hr)
CPV, TIC~141146667. The spectra show coherent sinusoidal Balmer
emission at twice to four times the star's equatorial velocity,
demonstrating the existence of extended clumps of corotating circumstellar
plasma --- a plasma torus.  Given that long-lived condensations of
cool ($10^4$ K) plasma can persist in the hot ($10^6$ K) coronae of
stars with a wide range of masses
\cite{CollierCameron1989,Townsend2005,Dunstone2006,Petit2013,Waugh2022,Daley-Yates2024},
these data support the idea that around the lowest-mass stars, such
condensations can be optically thick; the exact origin of the
optically thick material and its microphysical opacity remain unclear.
\end{abstract}

@@

%\subsection{Introduction}
M dwarfs, stars with masses below about half that of the Sun, are the
only type of star to offer near-term prospects for detecting the
atmospheres of rocky exoplanets with surface water
\cite{NAP26141}.  Investment with JWST has proceeded accordingly.  How
an M dwarf's evolution influences its planets---especially the
retention of their atmospheres--- has concurrently become a major
theme in exoplanet and stellar astrophysics.  Previous work has
established that most M dwarfs host close-in planets
\cite{Dressing2015} that are often subject to long
circumstellar disk lifetimes \cite{Ribas2015}, to large doses of UV
radiation \cite{France2013}, and to a high incidence of flares and
coronal mass ejections \cite{Gunther2020}.  However, despite excellent
work in these areas, the properties of the plasma and magnetospheric
environments that bathe young, close-in exoplanets have remained
challenging to quantify.

One glaring example of our current ignorance is the complex periodic
variables (CPVs).  While Figure~\ref{fig:lc} highlights the main object of
interest in this article, over one hundred analogous objects have
now been found by K2 and TESS 
\cite{Rebull2016,Stauffer2017,Rebull2018,Zhan2019,Rebull2020,Bouma2024}.
These CPVs are
phenomenologically identified based on their highly structured and
periodic optical light curves;
most are M dwarfs with rotation periods shorter than two days.
Within current sensitivity limits, none have primordial disks
\cite{Stauffer2017,Bouma2024}.
However, $\approx$3\% of stars a few million years old show this
behavior, and the observed fraction decreases to $\approx$0.3\%
by $\approx$150\,Myr \cite{Rebull2020}.

The two leading hypotheses to explain the CPVs are either that
transiting clumps of circumstellar material corotate with the star
\cite{Stauffer2017,Gunther2022,Bouma2024}, or that these stars represent
an extreme in naturally-occurring distributions of starspots or faculae
\cite{Koen2021}.  
% TODO: might want to plug Shapiro, Seager, ACC dark faculae...
While the main argument against a starspot-only explanation invokes the
timescales and amplitudes of the sharpest photometric features, no
independent evidence has yet been acquired for the presence of any
circumstellar material.  Geometrically, transiting clumps would imply an
occurrence a few to at most ten times the observed rate.  The clumps
could therefore exist around 10-30\% of M dwarfs during their early
lives.

\begin{figure}[!t]
  \centering
  \includegraphics[width=0.7\textwidth]{figures/f1.pdf}
  \captionsetup{labelformat=empty}
  \caption{{\bf Figure 1 (Movie):  TIC~141146667 is a complex periodic
  variable (CPV).}  The online movie,
  available
  \href{https://lgbouma.com/movies/TIC141146667_20250116.mp4}{here},
  shows a baseline of 5{,}784 cycles irregularly sampled over three
  years.  The TESS light curve is phased to the \periodhr\ hour period in
  groups of a few cycles per frame.  This is the period both of
  stellar rotation, and (we hypothesize) of corotating clumps of
  circumstellar material.  Raw data acquired with two minute sampling
  are in gray; black averages to 100 points per cycle.  Similar to
  other stars of this class, the sharp photometric features persist
  for tens to thousands of rotational cycles. }
  \label{fig:lc}
\end{figure}

The dearth of evidence for circumstellar material around CPVs is
surprising given that separate studies of young stars have, for
decades, reported that stellar coronae contain both hot ($10^6$ K) and
cool ($10^4$ K) plasma. In particular, time-series spectroscopy for
stars with a wide range of masses has shown periodic high-velocity
absorption and emission in Balmer lines such as H$\alpha$, interpreted
as long-lived, corotating clumps of cool plasma
\cite{CollierCameron1989,Donati2000,Dunstone2006,Skelly2008}.
Such clumps are forced into corotation by the magnetic field, and the
exact geometry of where the plasma can accumulate is dictated by the
field's topology.  For instance, a tilted dipole field tends
to yield an accumulation surface of a warped torus
\cite{Townsend2005}, whereas in the limit of a single strong field
line, accumulation occurs at the line's apex, furthest from the star
\cite{Waugh2022}.
To date, none of these spectroscopic variables have shown any
photometric anomalies \cite{Bouma2024}, leaving open the issue of
whether they are related to CPVs.
%A circumstantial argument that they might is that CPVs do respond to
%sudden magnetic field changes: the otherwise long-lived eclipse
%features often disappear following stellar flares
%\cite{Stauffer2017,Bouma2024}.

In this study, we present the first observations of corotating clumps of
cool plasma around a CPV.  We identified TIC~141146667 in previous work
\cite{Bouma2024} by searching the TESS two-minute data \cite{Ricker2015}
for stars showing at least three sharp, periodic dips per cycle.  We
selected the star for spectroscopy because its brightness and rapid
rotation enable efficiently studying small variations in the line profiles.  We
observed for five hours on 17~February~2024 (UT) using the High
Resolution Echelle Spectrometer (HIRES; \cite{vogt_hires_1994}) on the
Keck I 10m telescope.  TESS observed the star over a four week period
from 30~January~2024 to 26~February~2024 with a duty cycle of 77\%.
During the spectral observations, TESS was performing a data downlink;
TESS data collection resumed 12 hours (three rotation cycles) after the
spectra were acquired.  Extended Data Figure~\ref{fig:fulllc} shows the
detailed photometric behavior of the star before and after the exact
epoch of observation; the photometric morphology is similar before and
after the data gap.

@@

\begin{figure}[!tp]
  \centering
  \includegraphics[width=0.99\textwidth]{figures/f2.png}
  \captionsetup{labelformat=empty}
  \caption{{\bf Figure 2 (Movie):}  
  Hydrogen emission from circumstellar plasma orbiting TIC 141146667.
  {(\bf TODO)}For the best experience, please view the online movie
  available
  \href{https://lgbouma.com/movies/TIC141146667_sixpanel.mp4}{here}.
  {\bf Panel a:} TESS light curve from 5 February 2024 to 26 February
  2024 (UT) folded on the \periodhr\ hour period.  Black points are
  phase-averaged; gray are the raw data.
  {\bf Panel b:} Keck/HIRES H$\alpha$ spectra
  acquired on UT 17 February 2024.  The continuum is set to unity, and the
  darkest color is at twice the continuum to accentuate emission
  outside the line core ($|v/v_{\rm eq}|>1$, for $v_{\rm eq}$=130\,\kms).
  While emission in the line core originates in the stellar
  chromosphere, the sinusoidal emission features are most readily
  described by a warped plasma torus.
  {\bf Panel c:} Individual epochs of Panel b, visible in the
  online movie.  The dotted line shows the time-averaged spectrum,
  $f_{\langle t \rangle}$.
  {\bf Panel d:} As in Panel a, but overplotting the
  median-normalized H$\alpha$ light curve at $|v/v_{\rm eq}|<1$.
  {\bf Panel e:} As in Panel b, after subtracting the time-averaged
  spectrum. In addition to circumstellar emission, the line core shows
  absorption during the plasma clump transits.  The asymmetric stretch
  is set to match the dynamic range of the data.
  {\bf Panel f:} Individual epochs of Panel e, visible in the online
  movie.}
  \label{fig:spec}
\end{figure}

% photometry
Figure~\ref{fig:spec} shows the data from February 2024.  Over
timescales of years, CPVs maintain a fixed period while their
photometric morphology evolves.  TIC~141146667 indeed evolved relative
to the February 2022 discovery data \cite{Bouma2024}.  In February
2024, the average photometric signal showed a small brightening
over 45\% of the period, followed by a complex flux dip spanning 55\%
of the period.  This eclipse feature shows two to three local
photometric minima, and one to two local maxima.  Its W-shape is
suggestive of eclipse geometries seen in some forward models of warped
plasma tori \cite{Townsend2008}.

@@

% * keck/hires: behavior at v/veq<1
The behavior within the stellar H$\alpha$ line core, at $|\Delta v /
v_{\rm eq}|<1$, is more complex than outside it.  For stars of
this age and spectral type, one would expect emission in the line core
to be generated in the stellar chromosphere
and then modulated by any occulting material capable of absorbing or
emitting H$\alpha$ photons.  In Figure~\ref{fig:spec}e, the behavior from
$\phi$=0.4-1.2 has a simple interpretation: from $\phi$=0.4-0.9, a hot
region first gradually crosses the stellar line profile, followed from
$\phi$=0.7-1.2 by the transit of a cool region.  Phases $\phi$$<$0.4
show a mix of similar events, though the time sampling is
sufficiently coarse that the interpretation is less clear.  A final
exercise to quantify the behavior in the line core is shown in
Figure~\ref{fig:spec}d, where $f_{\rm H\alpha\ core}$ denotes the
summed flux at $|\Delta v / v_{\rm eq}|<1$.  This panel shows that
changes in the line core flux are usually correlated with the
broadband variability, except at $\phi$=0.5, during the transit of the
3.9\,$R_\star$ clump and the occultation of the lower-velocity clump.

@@

Spectra of magetically-active, rapidly rotating stars with a wide range
of masses have been known to exhibit both sinusoidal emission features
\cite{Donati2000,Townsend2005,Dunstone2006,Skelly2008} as well as
transient absorption features in their line cores
\cite{CollierCameron1989,CollierCameron1992,Cang2020} similar to
Figure~\ref{fig:spec}.  No such stars were previously known to show
complex light curves \cite{Bouma2024}.   One interpretation for
such spectroscopic variability comes from an analogy to quiescent solar
prominences, which are cool condensations of plasma in the solar corona
that can last days to weeks \cite{VialEngvold2015}.  In our Sun's
magnetosphere, these condensations fall back to the solar surface
because gravity is stronger than any magnetic tension or centrifugal force
capable of sustaining them.  However for stars with magnetospheric radii
$R_{\rm m}$ that exceed their corotation radii $R_{\rm c}$, the
effective potential experienced by a charged particle can have a local
minimum outside $R_{\rm c}$, enabling the material to accumulate over
much longer timescales \cite{Petit2013,Daley-Yates2024}.  Generally
speaking, such material need neither transit, nor be optically thick.

Our Keck/HIRES observations are the first reported time-series
spectra of a CPV, and they show that corotating circumstellar plasma clumps 
are present in at least one such star.
Characteristic densities and masses of these
%TODO check; quote N_H instead...
clumps are $n \sim 10^{10}$\,cm$^{-3}$ and $M \sim 10^{14}$\,kg (see
Supplementary Methods Section~\ref{subsec:model}), a similar density to
solar prominences, but ten to one hundred times more massive.  This
observation rules out a ``starspot-only'' origin scenario for CPVs,
\cite{Koen2021} since such scenarios have no means of explaining
spectroscopic emission beyond the stellar disk.  Similarly, scenarios in
which the circumstellar material is made only of dust are also ruled
out.  While dust may be present, to explain the H$\alpha$ emission the
clumps must include plasma with a significant population of hydrogen
atoms in the $n$=3 excited state.  While this plasma is undoubtedly
sculpted by the star's magnetic field, it could plausibly originate from
three sites: the star, an old and undetected disk, or outgassing rocky
bodies.  This latter possibility would render CPVs as extrasolar analogs
of the Jupiter-Io plasma torus (e.g.~Ref.~\cite{Bagenal1981}).

The other potential analog for the CPVs are the $\sigma$~Ori~E
variables, a rare subset of B stars with radiatively-driven winds that
accumulate into warped plasma tori \cite{Townsend2005,Townsend2008}.
These tori tend to have dense antipodal accumulations of plasma sculpted
by tilted-dipole magnetic fields, and the transits of these clumps are
thought to produce the observed broadband photometric variability
through bound-free scattering \cite{Townsend2005} and Thomson scattering
\cite{Berry2022}.  For $\sigma$~Ori~E and almost all of its analogs, the
result is light curves that appear ``simple'', resembling those of
eclipsing binaries \cite{Townsend2008}.  The two known exceptions,
HD~37776 and HD~64740, show complex light curves resembling CPVs
\cite{Mikulasek2020,Bouma2024} and have spectropolarimetric magnetic
field maps indicating strong contributions from higher-order magnetic
moments \cite{Kochukhov2011,Shultz2018}.  There are two implications:
first, the complexity of CPVs may be a direct consequence of magnetic
fields with highly multipolar contributions.  Second, CPVs can be a
source of astrophysical false positives in photometric searches for
eclipsing binaries and transiting exoplanets around young pre-main
sequence M dwarfs \cite{Johns-Krull2016,Bouma2020}.

@@

The material's composition -- either pure plasma, or else a dusty plasma
-- can be clarified by time-series optical and infrared
spectrophotometry.  While observations of CPVs in the optical have
previously suggested chromaticity consistent with dust
\cite{Tanimoto2020,Gunther2022,Koen2023}, a gray opacity source such as
electron scattering in a plasma transiting over starspots might also
produce chromatic features \cite{Rackham2018}.  The composition and size
distribution of any dust that is present could be determined
by measuring the extinction curve for one or more CPVs from
1-10\,$\mu$m.  Composition and size distributions similar to debris from
rocky bodies seen around white dwarfs \cite{Reach2009} would indicate a
rocky-body origin.  Compositions and sizes similar to the interstellar
medium would be
indicative of condensed dust in an M dwarf wind, similar to that formed
in the environments of more evolved stars \cite{Marigo2008}.

The role of the star's magnetic field could be better understood through
new observations, and new theory.  From the theoretical perspective,
there is an urgent need for rigid-field (magneto)-hydrodynamic modeling
to go beyond previous work \cite{Townsend2005,Townsend2008} and to
explore the effects of non-dipolar field contributions.  In particular,
dynamo simulations of fully-convective M dwarfs have suggested that
global-scale mean fields might be confined to a single hemisphere
\cite{Brown2020}; such fields could yield accumulation surfaces quite
different from those that have previously been explored
\cite{Townsend2008}.  Observationally, spectropolarimetry has the
potential to assess both the field strength and topology.  An
independent probe could also be connected the recent work
\cite{Kaur2024} showing that CPVs are variable radio emitters with
emission components that can be not only persistent, but also
short-lived and highly polarized.  In particular, detecting radio
emission produced by an electron cyclotron maser instability
\cite{Callingham2021} would provide a measurement of the field
strength at the site of the emitting region.

It is currently unclear what, if any, relationship CPVs have to the
close-in rocky exoplanets that exist around most M dwarfs
\cite{Dressing2015}.  However, 0.3-3\% of young M dwarfs show the CPV
phenomenon \cite{Rebull2020}, and our data show that the complex
photometric features occur when clumps of circumstellar material is
transiting the star.  The implied geometric correction suggests that
an appreciable minority (10-30\%) of young M dwarfs -- the rapidly
rotating ones with centrifugal magnetospheres -- host similar
circumstellar environments to the CPVs.

@@

\subsection{Spectroscopic Variability}\phantom{+}

>>>>

\begin{abstract}
\normalfont
Roughly one percent of red dwarfs younger than 100 million years
show structured, periodic optical light curves suggestive of
transiting opaque material that corotates with the star
\cite{Rebull2016,Stauffer2017,Rebull2018,Bouma2024}.  The
composition, origin, and even the existence of this material are
uncertain. The main alternative hypothesis is that these complex
periodic variables (CPVs) are explained by complex distributions of
bright or dark regions on the stellar surfaces
\cite{Koen2021}.  Here, we present time-series spectroscopy and
photometry of a rapidly-rotating ($P$=3.9\,hr)
CPV, TIC~141146667. The spectra show coherent sinusoidal Balmer
emission at twice to four times the star's equatorial velocity,
demonstrating the existence of corotating clumps of circumstellar
plasma --- a plasma torus.  Given that long-lived condensations of
cool ($10^4$ K) plasma can persist in the hot ($10^6$ K) coronae of
stars with a wide range of masses
\cite{CollierCameron1989,Townsend2005,Dunstone2006,Petit2013,Waugh2022,Daley-Yates2024},
these data support the idea that around the lowest-mass stars, such
condensations can sustain an optically thick component; the 
origin of the optically thick material and its microphysical opacity
remain unclear.
\end{abstract}

@@

%\subsection{Introduction}
M dwarfs, stars with masses below about half that of the Sun, are the
only type of star to offer near-term prospects for detecting the
atmospheres of rocky exoplanets with surface water \cite{NAP26141}.
Investment with JWST has been proceeding accordingly
\cite{Redfield2024,TRAPPIST1JWSTCommunityInitiative2024}.  How an M
dwarf's evolution influences its planets---especially the retention of
their atmospheres--- has concurrently become a major theme in
exoplanet and stellar astrophysics.  Previous work has established
that most M dwarfs host close-in planets \cite{Dressing2015} that are
on average subject to long circumstellar disk lifetimes \cite{Ribas2015},
to large doses of UV radiation \cite{France2013}, and to a high
incidence of flares and coronal mass ejections \cite{Gunther2020}.
However, despite excellent work in these areas, the properties of the
plasma and magnetospheric environments that bathe young, close-in
exoplanets have remained challenging to quantify.

One glaring example of our current ignorance is the complex periodic
variables (CPVs).  While Figure~\ref{fig:lc} highlights the main object of
interest in this article, over one hundred analogous objects have
now been found by K2 and TESS 
\cite{Rebull2016,Stauffer2017,Rebull2018,Zhan2019,Rebull2020,Bouma2024}.
These CPVs are
phenomenologically identified based on their structured, periodic
optical light curves;
most are M dwarfs with rotation periods shorter than two days.
Within current sensitivity limits, none have primordial disks
\cite{Stauffer2017,Bouma2024}.
However, $\approx$3\% of stars a few million years old show this
behavior, an  observed fraction which decreases to $\approx$0.3\%
by $\approx$150\,Myr \cite{Rebull2020}.

The two leading hypotheses to explain the CPVs are either that
transiting clumps of circumstellar material corotate with the star
\cite{Stauffer2017,Gunther2022,Bouma2024}, or that these stars represent
an extreme in naturally-occurring distributions of starspots or faculae
\cite{Koen2021}.  
The main argument against a starspot-only explanation invokes the
timescales and amplitudes of the sharpest photometric features.
However, no independent evidence has yet been acquired for the
presence of any circumstellar material.  Geometrically, transiting
clumps would imply an occurrence a few to at most ten times the
observed rate.  The clumps could therefore exist around 10-30\% of M
dwarfs during their early lives.

\begin{figure}[!t]
  \centering
  \includegraphics[width=0.7\textwidth]{figures/f1.pdf}
  \captionsetup{labelformat=empty}
  \caption{{\bf Figure 1 (Movie):  TIC~141146667 is a complex periodic
  variable (CPV).}  The online movie,
  available
  \href{https://lgbouma.com/movies/TIC141146667_20250116.mp4}{here},
  covers a baseline of 5{,}784 cycles irregularly sampled over three
  years.  The TESS light curve is phased to the \periodhr\ hour period in
  groups of a few cycles per frame.  This is the period both of
  stellar rotation, and (we hypothesize) of corotating clumps of
  circumstellar material.  Raw data acquired with two minute sampling
  are in gray; black averages to 100 points per cycle.  Similar to
  other stars of this class, the sharp photometric features persist
  for tens to thousands of rotational cycles. }
  \label{fig:lc}
\end{figure}

The dearth of evidence for circumstellar material around CPVs is
surprising given that separate studies of young stars have, for
decades, reported that stellar coronae contain both hot ($10^6$ K) and
cool ($10^4$ K) plasma. In particular, time-series spectra of
stars with a wide range of masses have shown periodic high-velocity
absorption and emission in Balmer lines such as H$\alpha$, interpreted
as long-lived, corotating clumps of cool plasma
\cite{CollierCameron1989,CollierCameron1992,Barnes2000,Donati2000,Dunstone2006,Skelly2008,Leitzinger2016,Cang2021}.
Such clumps are thought to be forced into corotation by the magnetic field, and the
exact geometry of where the plasma can accumulate is dictated by the
field's topology.  For instance, a tilted dipole field tends
to yield an accumulation surface of a warped torus
\cite{Townsend2005}, whereas in the limit of a single strong field
line, accumulation occurs at the line's apex, furthest from the star
\cite{Waugh2022}.
To date, none of these spectroscopic variables have shown any
photometric anomalies \cite{Bouma2024}, leaving open the issue of
whether they are related to CPVs.
%A circumstantial argument that they might is that CPVs do respond to
%sudden magnetic field changes: the otherwise long-lived eclipse
%features often disappear following stellar flares
%\cite{Stauffer2017,Bouma2024}.

In this study, we present the first detection of corotating clumps of
cool plasma around a CPV.  We identified TIC~141146667 in previous work
\cite{Bouma2024} by searching the TESS two-minute data \cite{Ricker2015}
for stars showing at least three sharp, periodic dips per cycle.  We
selected the star for spectroscopy because its brightness and 
rotation rate enable efficiently mapping out any line profile variability.  We
observed for five hours on 17~February~2024 (UT) using the High
Resolution Echelle Spectrometer (HIRES \cite{vogt_hires_1994}) on the
Keck I 10m telescope.  TESS observed the star over a four week interval
from 30~January~2024 to 26~February~2024 with a duty cycle of 77\%.
During the spectral observations, TESS was performing a data downlink;
TESS data collection resumed 12 hours (three rotation cycles) after the
spectra were acquired.  Extended Data Figure~\ref{fig:fulllc} shows the
detailed photometric behavior of the star before and after the exact
epoch of observation; the general photometric morphology is similar before and
after the data gap.

@@

\begin{figure}[!tp]
  \centering
  \includegraphics[width=0.99\textwidth]{figures/f2.png}
  \captionsetup{labelformat=empty}
  \caption{{\bf Figure 2 (Movie):}  
  Hydrogen emission from circumstellar plasma orbiting TIC 141146667.
  The online movie, available
  \href{https://lgbouma.com/movies/TIC141146667_sixpanel.mp4}{here},
  shows the spectral evolution over five hours.
  {\bf Panel a:} Average TESS light curve from 5 February 2024 to 26
  February 2024 (UT) folded on the \periodhr\ hour period.  Black
  points are phase-averaged; gray are the raw data.
  {\bf Panel b:} Keck/HIRES H$\alpha$ spectra acquired on UT 17
  February 2024.  The continuum is set to unity, and the darkest color
  is at twice the continuum to accentuate emission outside the line
  core ($|v/v_{\rm eq}|>1$, for $v_{\rm eq}$=130\,\kms).  While
  emission in the line core originates in the star's chromosphere,
  the sinusoidal emission features are most readily described by a
  warped plasma torus.
  {\bf Panel c:} Individual epochs of Panel b, visible in the
  online movie.  The dotted line shows the time-averaged spectrum,
  $f_{\langle t \rangle}$.
  {\bf Panel d:} As in Panel a, but overplotting the
  median-normalized H$\alpha$ light curve at $|v/v_{\rm eq}|<1$.
  {\bf Panel e:} As in Panel b, after subtracting the time-averaged
  spectrum. In addition to circumstellar emission, the line core shows
  absorption during the plasma clump transits.  The asymmetric stretch
  is set to match the dynamic range of the data.
  {\bf Panel f:} Individual epochs of Panel e, visible in the online
  movie.}
  \label{fig:spec}
\end{figure}

% photometry
Figure~\ref{fig:spec} shows the TESS and HIRES data from February
2024.  Over timescales of years, CPVs maintain a fixed period while
their photometric morphology evolves.  TIC~141146667 indeed evolved
relative to the February 2022 discovery data (see Figure~\ref{fig:lc}.
In February 2024, the average photometric signal showed a small
brightening over 45\% of the period, followed by a complex flux dip
spanning 55\% of the period.  This eclipse feature shows two to three
local photometric minima, and one to two local maxima.  Its W-shape is
suggestive of eclipse geometries seen in some forward models of warped
plasma tori \cite{Townsend2008}.

@@

% * keck/hires: behavior at v/veq<1
The behavior within the stellar H$\alpha$ line core, at $|\Delta v /
v_{\rm eq}|<1$, is more complex than outside it.  For stars of
this age and spectral type, one would expect emission in the line core
to be generated in the star's chromosphere
and then modulated by any occulting material capable of absorbing or
emitting H$\alpha$ photons.  In Figure~\ref{fig:spec}e, the behavior from
$\phi$=0.4-1.2 has a simple interpretation: from $\phi$=0.4-0.9, a hot
region first gradually crosses the stellar line profile, followed from
$\phi$=0.7-1.2 by the transit of a cool region.  Phases $\phi$$<$0.4
show a mix of similar events, though the time sampling is
sufficiently coarse that the interpretation is less clear.  A final
exercise to quantify the behavior in the line core is shown in
Figure~\ref{fig:spec}d, where $f_{\rm H\alpha\ core}$ denotes the
summed flux at $|\Delta v / v_{\rm eq}|<1$.  This panel shows that
changes in the line core flux are usually correlated with the
broadband variability, except at $\phi$=0.5, during the transit of the
3.9\,$R_\star$ clump and the occultation of the lower-velocity clump.

@@

Spectra of magetically-active, rapidly rotating stars with a wide range
of masses have been known to exhibit both sinusoidal emission features
\cite{Donati2000,Townsend2005,Dunstone2006,Skelly2008} as well as
transient absorption features in their line cores
\cite{CollierCameron1989,CollierCameron1992,Cang2020} similar to
Figure~\ref{fig:spec}.  No such stars were previously known to show
complex light curves \cite{Bouma2024}.   One interpretation for
such spectroscopic variability comes from an analogy to quiescent solar
prominences, which are cool condensations of plasma in the solar corona
that can last days to weeks \cite{VialEngvold2015}.  These
condensations fall back to the solar surface because gravity is
stronger than any magnetic tension or centrifugal force capable of
sustaining them.  However for stars with magnetospheric radii $R_{\rm
m}$ that exceed their corotation radii $R_{\rm c}$, the effective
potential experienced by a charged particle can have a local minimum
outside $R_{\rm c}$, enabling the material to accumulate over much
longer timescales \cite{Petit2013,Daley-Yates2024}.  Generally
speaking, such material need neither transit, nor be optically thick.
For instance, the H$\alpha$ emission clump at 3.9\,$R_\star$ shows no 
photometric signal in the February 2024 photometry.

Our Keck/HIRES observations are the first reported time-series
spectra of a CPV, and they show that corotating circumstellar plasma clumps 
are present in at least one such star.
Based on the observed H$\alpha$ luminosity, the characteristic
densities and masses of the gaseous component of these clumps are
$n_{\rm H} \sim 10^{11}$-$10^{12}$\,cm$^{-3}$ and $M_{\rm gas} \sim 10^{17}$\,g
(see Section~\ref{subsec:gas}).
Any dust, if present, is independently constrained to have a total
mass $10^{15}\,{\rm g} < M_{\rm dust} < 10^{17}\,{\rm g}$ (see
Section~\ref{subsec:dust}).
This
observation rules out a ``starspot-only'' origin scenario for CPVs,
\cite{Koen2021} because such scenarios have no means of explaining
spectroscopic emission beyond the stellar disk.  Similarly, scenarios in
which the circumstellar material is made only of dust are also ruled
out.  While dust may be present, to explain the H$\alpha$ emission the
clumps must include plasma with a significant population of hydrogen
atoms in the $n$=3 excited state.  While this plasma is undoubtedly
sculpted by the star's magnetic field, it could plausibly originate from
three sites: the star, an old and undetected disk, or outgassing rocky
bodies.  This latter possibility would render CPVs as extrasolar analogs
of the Jupiter-Io plasma torus (e.g.~Ref.~\cite{Bagenal1981}).

The other potential analog for the CPVs are the $\sigma$~Ori~E
variables, a rare subset of B stars with radiatively-driven winds that
accumulate into warped plasma tori \cite{Townsend2005,Townsend2008}.
These tori tend to have dense antipodal accumulations of plasma sculpted
by tilted-dipole magnetic fields, and the transits of these clumps are
thought to produce the observed broadband photometric variability
through bound-free scattering \cite{Townsend2005} and Thomson scattering
\cite{Berry2022}.  For $\sigma$~Ori~E and almost all of its analogs, the
result is light curves that appear ``simple'', resembling those of
eclipsing binaries, and ``S-wave'' H$\alpha$ spectra similar to those
in Figure~\ref{fig:spec} \cite{Townsend2005,Townsend2008}.  The two known exceptions,
HD~37776 and HD~64740, show complex light curves resembling CPVs
\cite{Mikulasek2020,Bouma2024} and have spectropolarimetric magnetic
field maps indicating strong contributions from higher-order magnetic
moments \cite{Kochukhov2011,Shultz2018}.  There are two implications:
first, the complexity of CPVs may be a direct consequence of magnetic
fields with highly multipolar contributions.  Second, CPVs can be a
source of astrophysical false positives in photometric classification
of eclipsing binaries and transiting exoplanets around young pre-main
sequence M dwarfs \cite{Johns-Krull2016,Bouma2020}.

@@

The material's composition -- either pure plasma, or else a dusty plasma
-- can be clarified by time-series optical and infrared
spectrophotometry.  While observations of CPVs in the optical have
previously suggested chromaticity consistent with dust
\cite{Tanimoto2020,Gunther2022,Koen2023}, a gray opacity source such as
electron scattering in a plasma transiting over starspots might also
produce chromatic features \cite{Rackham2018}.  The composition and size
distribution of any dust that is present could be determined
by measuring the extinction curve for one or more CPVs from
1-10\,$\mu$m.  Composition and size distributions similar to debris from
rocky bodies seen around white dwarfs \cite{Reach2009} would indicate a
rocky-body origin.  Compositions and sizes similar to the interstellar
medium could be indicative of condensed dust in an M dwarf wind,
similar to that formed in the environments of more evolved stars
\cite{Marigo2008}.

The role of the star's magnetic field could be better understood through
new observations, and new theory.  From the theoretical perspective,
there is an urgent need for rigid-field (magneto)-hydrodynamic modeling
to go beyond previous work \cite{Townsend2005,Townsend2008} and to
explore the effects of non-dipolar field contributions.  In particular,
dynamo simulations of fully-convective M dwarfs have suggested that
global-scale mean fields might be confined to a single hemisphere
\cite{Brown2020}; such fields could yield accumulation surfaces quite
different from those that have previously been explored
\cite{Townsend2008}.  Observationally, spectropolarimetry has the
potential to assess both the field strength and topology.  An
independent probe could also be connected the recent work
\cite{Kaur2024} showing that CPVs are variable radio emitters with
emission components that can be not only persistent, but also
short-lived and highly polarized.  In particular, detecting radio
emission produced by an electron cyclotron maser instability
(e.g. \cite{Callingham2021}) would provide a measurement of the field
strength at the site of the emitting region.

It is currently unclear what, if any, relationship CPVs have to the
close-in rocky exoplanets that exist around most M dwarfs
\cite{Dressing2015}.  However, 0.3-3\% of young M dwarfs show the CPV
phenomenon \cite{Rebull2020}, and our data show that the complex
photometric features occur when clumps of circumstellar material 
transit the star.  The implied geometric correction suggests that
an appreciable minority (10-30\%) of young M dwarfs -- the rapidly
rotating ones with centrifugal magnetospheres -- host similar
circumstellar environments to the CPVs.

@@

\subsection{Spectroscopic Variability}\phantom{+}
\label{subsec:specvar}
